---
title: "Exercise 3"
author: "Adam Sulak, Onder Akacik, Arda Ergin"
date: "2024-02-22"
output:
  pdf_document: default
  html_document: default
---
# Exercise 3
```{r}
data <- read.table("cream.txt", header = TRUE)
```

## Question 3a)
```{r}
# Factorizing the variables
data$batch <- factor(data$batch)
data$position <- factor(data$position)
data$starter <- factor(data$starter)

model <- stats::lm(acidity ~ batch + position + starter, data = data)
# summary(model)
```

The comparison category for the model is `starter = 1`. Hence, the `starter2` table essentially gives us whether there is a "*significant difference between the effects of starter 1 and starter 2 on acidity*". And the results of the analysis shows that, there is no significant difference between the two $t(12) = -0.321$, $p = 0.7534$, on their effect on acidity. This is also the case for starter 2, 3, and 5.


## Question 3b)
```{r}
anova(model)
```

The results of the ANOVA show that `position` variable does not have a significant effect on acidity, $F(4,12) = 1.08$, $p = 0.41$. And the `summary()` above also had given us that there is no significant differences between any categories with `position1`. Therefore, this variable can be removed from the model.
```{r}
model_reduced <- stats::lm(acidity ~ batch + starter, data = data)
anova(model_reduced)
# summary(model_reduced)
```

In this reduced model, we see that the `batch` variable have a significant effect on acidity, $F(4,16) = 8.44$, $p < .001$, as well as the `starter` variable, $F(4,16) = 19.83$, $p < .001$. When we further investigate the model summary, we can see that `starter4` significantly differs from `starter1` on its effect on acidity, $t(16) = 5.96$, $p < .001$. The other starters do not have a significant differ from `starter1`.


## Question 3c)
In our situation, **Friedman Rank Sum Test** could be a reasonable alternative. This is because batches can essentially be considered as "repeated-measures", which is the applicable situation for the Friedman Test. As a non-parametric test, it offers an alternative to repeated-measures ANOVA, and violations of the normality assumption do not bother it as it bothers ANOVA.

```{r}
stats::friedman.test(acidity ~ starter | batch, data = data)
```

Similar to the ANOVA results, the results of the Friedman Rank Sum Test also shows that there is a statistically significant difference in acidity, $\chi^2(4) = 13.21$, $p = .01$.


## Question 3d)
```{r}
lme4::lmer(acidity ~ starter + (1|batch), data = data)
# summary(lmerTest::lmer(acidity ~ starter + (1|batch), data = data))
# anova(lmerTest::lmer(acidity ~ starter + (1|batch), data = data))
```

Here, in this model, we register `batch` as a "random effect". Here, through `(1|batch)`, we specify that we are allowing the intercept to vary across batches, by including a random intercept for each level of batch (as we indicate with `1`). This, in return, allows us to account for variability that is not of main interest (as, our main interest is the `starter`) but should be accounted for in the model.

We can observe that $\sigma = 0.91$


The results show that the standard deviation of the residuals is $0.90$, which show the variability within batches that is not explained by the model, and we can see that there is a certain variability that we are accounting through the random effects.

The Fixed Effects show the effects of different starters on acidity. Compared to Starter 1 Starters 2, 3, and 5 lead to an estimated decrease in the acidity by units of $-0.15$, $-0.98$, and $-0.48$ respectively. On the other hand, Starter 4, compared to Starter 1, leads to an estimated increase on acidity by unit of $2.81$.

We can further test for significance with `lmerTest::lmer()`. If we conduct an ANOVA on the model, we see that the `starter` variable has a significant effect overall, $F(4,16) = 19.83$, $p < .001$ We also see that while the differences of Starter 2, 3, and 5 do not have a significant difference from Starter 1, Starter 4 seems to lead to a significantly higher acidity level, $t(16) = 5.96$, $p < .001$. 

Interestingly, the F-value for the starter variable and the t-value for Starter 1 vs. Starter 4 comparison is the same as the fixed model, and in both models, the significance are the same.



