---
title: "Exercise 3"
author: "Adam Sulak, Onder Akacik, Arda Ergin"
date: "2024-02-22"
output:
  pdf_document: default
  html_document: default
---
# Exercise 3

The file `cream.txt` contains data on an experiment to produce sour cream. 

```{r}
data <- read.table("cream.txt", header = TRUE)
```


## Question 3a)
```{r}
# Factorizing the variables
data$batch <- factor(data$batch)
data$position <- factor(data$position)
data$starter <- factor(data$starter)

# Linear Modelling
model <- stats::lm(
  acidity ~ batch + position + starter, data = data)
summary(model)
```

**Answer**: 
The comparison category for the model is `starter = 1`. 

Hence, the `starter2` table essentially gives us whether there is a "*significant difference between the effects of starter 1 and starter 2 on acidity*". 

And the results of the analysis shows that, there is no significant difference between the two $t(12) = -0.321$, $p = 0.7534$.



## Question 1b)
b)  Recall that the main interest is in the effect of starter on acidity; factors `position` and `batch` represent the block variables. Remove insignificant block variable(s) if there are such, and perform an ANOVA for the resulting “fixed effects” model. Which starter(s) lead to significantly different acidity? Motivate your answer.

```{r}
anova(model)
```

We see that, overall, `position` variable is insignificant. And the `summary()` above also gives us that there is no significant differences between any categories with `position1`. 

Therefore, it can be removed from the model.
```{r}
model_reduced <- stats::lm(
  acidity ~ batch + starter, 
  data = data)

anova(model_reduced)
summary(model_reduced)
```


## Question C)
```{r}
stats::friedman.test(
  acidity ~ starter | batch,
  data = data)
```


## Question D)
```{r}
model_mixed <- lme4::lmer(
  acidity ~ starter + (1|batch), 
  data = data)
summary(model_mixed)
anova(model_mixed)
```




